---
title: "Exercise 3"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Exercise 3-1

```{r}
library(tidyverse)
library(GGally)
library(lubridate)
dt_clean = read_rds("data/dt_clean.rds")
dt_clean_week = dt_clean %>%
  filter(date(datetime) < "2004-05-05")

variables = c("humidity", "humid_temp","hamabot","hamatop")
dt_clean_week[,variables] %>% 
ggpairs()
```

We decided to plot scatterplots using time period of 1 week. We found this to be reasonable as some nodes began failing afterwards. We notice similar correlation structure between the weeks. We observe that Incident and Reflected PAR have significant positive correlation (~0.475). This is expected as both measures relate to sunlight. Additionally, humidity and temperature are strongly, negatively correlated (~0.671), which makes sense because according to the paper, in California, warm days are dry and colder days are humid.

### Exercise 3-2

```{r}
variables = c("humidity", "humid_temp","hamabot","hamatop", "voltage", "height", "dist")
dt_clean_week[,variables] %>% 
ggpairs()
```

We added voltage, height and distance to check if they have correlation with incident PAR. We notice that in addition to the positive association between, Incident and Reflected PAR, there is a positive association between Incident PAR and height ~0.305. This makes sense because the higher up the node, the more the exposure to sunlight.

### Exercise 3-3

### Exercise 3-4

```{r}
variables = c("humidity", "humid_temp","hamabot","hamatop","height","dist")
pca.in = dt_clean[,variables] %>% filter(!is.na(height))
pr.out <- prcomp(pca.in, scale = T)
```

```{r}
pve = 100 * pr.out$sdev^2 / sum(pr.out$sdev ^2)
par (mfrow = c (1 , 2))
plot(pve,type = "o",ylab ="PVE",xlab = "Principal Component", col ="blue")
plot (cumsum (pve), type = "o",ylab ="Cumulative PVE", xlab =" Principal Component ",col=" brown3")
pr.out$rotation
```

Yes, we can represent data in lower dimension. Looking at cumulative PVE we see that including 3 or 4 principal components allows us to retain around 80%, 90% of variance respectively while eliminating 3,4 dimensions with less variance.

### Exercise 4-1

### Exercise 4-2

### Exercise 5-a

hist(dt_clean$hamatop)

```{r}
dt_clean %>% 
select(hamatop) %>% 
mutate(log_hamatop = log(hamatop)) %>% 
ggplot(aes(x=log_hamatop)) + geom_histogram(binwidth = 0.5) 

dt_clean %>% 
select(hamabot) %>% 
mutate(log_hamabot = log(hamabot)) %>% 
ggplot(aes(x=log_hamabot)) + geom_histogram(binwidth=0.4)
```



### Exercise 5-b

From 3-c, we can see that incident par and reflected par vary along the different heights whereas humidity and temperature are more consistent vs height.

From 3-d, we can see that incident par and reflected par from the mean depending on height. higher heights tend to have higher indicident and reflected par than lower heights.

